Software architectures to integrate workflow engines in science gateways

### 2016-04-25 paper.tex
fixed structure of the paper
Intro - 2.1 minor edits
Minor edits and comments abstract to up to 2.1 (excluded).

\author[mcgill,creatis]{Tristan Glatard}
\author[mcgill]{Marc-Etienne Rousseau}
\author[creatis]{Sorina Camarasu-Pop}

@@

\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}

\begin{abstract}
  We investigate software architectures to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract software components
  and interactions in 6 architectures that we use to classify 18
  systems. We compare these architectures using metrics measuring
  integration effort, robustness, extensibility, scalability and other
  specific features. Results show the pros and cons of the different
  architectures: tight integration and sub-tasking require the least
  amount of integration effort and are also the most robust, most
  architectures perform the same in terms of extensibility, the pool
  model is the most scalable and meta-workflows are only available in
  nested workflows and workflow import. These results provide insights
  to software architects on how to integrate science gateways and
  workflow engines.
\end{abstract}

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to handle high-throughput
task executions. For instance, FSL can launch tasks on SGE through its
fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF and Condor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now widely shared across neuroimaging groups
world-wide. These
domain-specific engines represent a tremendous opportunity for science
gateways to leverage existing tools and applications. They nicely
complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science and
domain-specific engines. Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave it to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources in a community and to
provide performance through facilitated access to storage and
computing power. They are usually accessible through a web interface
that helps users manage access rights, data transfers, task execution
and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

Robustness is a key part of science gateway user experience. In this
respect, the multiplicity of software components and interactions
involved in architectures is a challenge as it often leads to complex
systems that are therefore prone to failure. To offer good user
experience, science gateways should remain as transparent as possible,
which requires robustness.

@@

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including nine national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As September 2015, CBRAIN has served over 370 users across 50
cities in 19 countries who produced several high-impact results in the
field of neurosciences.

@@

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, and the software
services used to access these resources. Infrastructures can be
computing or file servers, databases, clusters, grids or clouds. Some
workflow engines and science gateways may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes, the availability of a
global task meta-scheduler, the presence of a file catalog, etc. In
the analysis presented below, such specific functions are not
discussed. Instead, we consider the infrastructure as an abstract
system that can execute tasks and store data regardless of the
enabling mechanisms.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}.

@@

Software interactions among these components are represented with
arrows.  \emph{Abstract} interactions are specific types of
interactions that may be implemented using different components and
interactions. They are represented with dotted arrows and labeled with
\texttt{*}. An architecture that has an abstract interaction is an
abstract architecture that can be instantiated in various ways.

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g. unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.

>>>>

\author[mcgill,creatis]{Tristan Glatard}
\author[mcgill]{Marc-\'Etienne Rousseau}
\author[creatis]{Sorina Camarasu-Pop}

@@

\note{Marc-Etienne}{Do we want to use MCIN instead of BIC?}
\address[mcgill]{McConnell Brain Imaging Centre, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montreal, QC, Canada.}

\begin{abstract}
  We investigate software architectures commonly used to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract software components
  and interactions in 6 architectures that we use to classify 18
  systems. We compare these architectures using metrics measuring
  integration effort, robustness, extensibility, scalability and other
  specific features. Results show the pros and cons of the different
  architectures: tight integration and sub-tasking require the least
  amount of integration effort and are also the most robust, most
  architectures perform the same in terms of extensibility, the pool
  model is the most scalable and meta-workflows are only available in
  nested workflows and workflow import. These results provide insights
  for software architects on how to integrate science gateways and
  workflow engines.
\end{abstract}

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to handle high-throughput
task executions. For instance, FSL can launch tasks on SGE through its
fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF and Condor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These
domain-specific engines represent a tremendous opportunity for science
gateways to leverage existing tools and applications. They nicely
complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science and
domain-specific engines. Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources in a community and to
provide increased performance and capacity through facilitated access to storage and
computing power. They are often accessible through a web interface
that helps users manage access rights, data transfers, task execution
and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

Robustness is a key part of science gateway user experience. In this
respect, the multiplicity of software components and interactions
involved in architectures is a challenge as it often leads to complex
systems that are therefore prone to failure. To offer good user
experience, science gateways should remain as transparent as possible,
which requires robustness.
\note{Marc-Etienne}{I find that something is weak in this last
paragraph. Robustness will help with transparency, sure, but it seems only a small facet.
Also, the text seem to timidly favor architectures with less components/interactions
without much explanation... Complex systems can be robust, but at greater costs...
Here is one:
"Architectures showing reduced system complexity will greatly facilitate the implementation
effort towards robust interactions which, in turn, have a positive impact on characteristics
such as gateway predictability, transparence, general reliability and results traceability and
reproducibility; characteristics that are also key to a good user experience."}

@@

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including a network of national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As February 2016, CBRAIN has served over 420 users across 50
cities in 22 countries and has a national computing allocation of 660 CPU years
per annum.

@@

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, as well as the software
services used to access these resources. Infrastructures can be composed of
computing or file servers, databases, clusters, grids or clouds. Some
workflow engines and science gateways may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes, the availability of a
global task meta-scheduler, the presence of a file catalog, etc. In
the analysis presented below, such specific functions are not
discussed. Instead, we consider the infrastructure as an abstract
system that can execute tasks and store data regardless of the
enabling mechanisms.

@@

Architectures to integrate workflow engines in science gateways are
diagrammed in Figure~\ref{fig:architectures} using the graphical
notations shown in Figure~\ref{fig:notations}.
\note{Marc-e}{A few things bother me a bit with the notation fig., making it harder to assimilate
1- The interactions are padded in the middle of other stuff (Administrator, Data)
2- The Software component example says "Science gateway", should it not be empty?
3- The red arrow to explain the use of red; it looks like a very specific type 
of interactions instead of the desired "any interactions or components can be red meaning...",
especially with 'a' and 'd' which the reader does not know the meaning yet.
Maybe a little red arrow and square with "Red used when specific to workflow execution"?
I don't have a great solution...
4- Order in text jumps all over the fig notations. Explanations are given for, in order:
Components, workflow engine, interactions, abstract interactions, red color, tasks, data (no Administrator?)...
Could re-ordering the fig fix this? A figure figure legend could go a long way too (are there some in this journal?)
}

@@

Software interactions among these components are represented with labeled
arrows.  \emph{Abstract} interactions are specific types of
interactions that may be implemented using different components and
interactions. They are represented with dotted arrows and labeled with
\texttt{*}. An architecture that has an abstract interaction is an
abstract architecture that can be instantiated in various ways.

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g. unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.
